**Bayesian statistics for longitudinal studies in biomedical research**

### 2020-11-11 Manuscripts/Manuscript_AM_v1.Rmd
Updated manuscript

A longitudinal study is defined as that which is designed to repeatedly measure a variable of interest in a group (or groups) of subjects. In biomedical research, this type of study arises when the investigator intends to observe the evolution of the effect of a certain treatment across time, rather than analyzing it at a single time point (a cross-sectional study). Clinical examples of this  approach in biomedical research include studies on breast and neck cancer[@sio2016;@kamstra2015]; in the first case, weekly measurements of skin toxicities in  patients with radiation-induced dermatitis were taken for up to 8 weeks; whereas in the latter mouth opening was assessed at 6,12, 18, 24 and 36 months after radiotherapy (RT). Longitudinal studies have used also to measure tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression[@ritter2001;@roth2017], and cell metabolism[@jones2018;@skala2010]. From a statistical standpoint, a longitudinal study presents advantages over a cross-sectional approach:it requires a lower number of subjects to reach a certain statistical power, and besides it being able to track the previously mentioned time-effect evolution on a group-by-group basis, it allows to determine the variability of the response within subjects [@guo2013; @fitzmaurice2012]. In other words,a longitudinal study permits to quantify how the variable changes within each subject across time.

@@

However, constant correlation is frequently unjustified as its value tends to diminish between measures when the time interval between them increases[@ugrinowitsch2004], and the violation of this assumption increases the false positivity rate [@lane2016].
Moreover, it is unlikely that complete observations in all subjects are obtained in a biomedical study due to reasons that can be specific to different situations. In a clinical trial, voluntary withdrawal from one or multiple patients can occur, whereas attrition in animals due to injury or weight loss can occur in preclinical experiments, and in both cases it is possible that unexpected complications with equipment or supplies arise, preventing the researcher from collecting measurements at a certain time point.

When the aforementioned issues arise, rm-ANOVA requires to exclude all subjects with missing observations from the analysis, thereby increasing costs for the study if the desired statistical power is not met with the remaining observations as it makes necessary to enroll more subjects; and  raising the possibility that the rejection of those partial observations limits the demonstration of significant differences between groups. Additionally, rm-ANOVA uses a _post hoc_ analysis to assess the relevance between the measured response in different groups. This analysis is based on multiple repeated comparisons to estimate a _p-value_, a metric that is widely used as a measure of significance. Because the p-value is highly variable [@halsey2015;@nuzzo2014], multiple comparisons can inflate the false positivity rate [@liu2010],consequently biasing the conclusions of the study.

@@

Additionally, another assumption is made in both rm-ANOVA and LMEMs models, where a linear relationship is expected between the observed response and the covariates across the study [@pinheiro2006].This common assumption to both rm-ANOVA and LMEMs causes the models to be restrictive in their inferences when used in longitudinal data that does not follow a linear trend. In biomedical research, this particular behavior in longitudinal has been reported in studies of tumor response to radio/chemotherapy in preclinical and clinical settings[@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994]. This studies have shown that that the collected signal does not follow a linear trend over time, and presents high variability at different time points, making the estimations of a LMEM or rm-ANOVA model inconsistent with the pattern of the observed variations. Additionally,  although it is possible that a post hoc analysis is able to find “significance”( _p-value_<0.05) by using multiple comparisons between the model terms, this estimator would be inherently biased because of the lack of fit between the information and the model.

As  the “frequentist” rm-ANOVA and the more advanced LMEM approach are both limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed distribution) to determine the fit of the model while enabling inferences that are both adequate and consistent from a statistical perspective.

Generalized additive models (GAMs) are a subset of generalized linear models that use _smooth functions_ (hencefort _smooths_) to estimate the parameters of a model. They have been used in palaeolimnology, ecology and clinical studies to model longitudinal data[@woolway2016;@hefley2017;@ko2007]. Briefly, GAMs use a combination of multiple functions (basis functions)to construct the smooths of the model [@wood2017]. Their main advantage over LMEMs and rm-ANOVA is that the model specification is directed by the _smooths_ rather than by a parametric relationship. This allows a consistent fit of the model with the data, and estimations of significance using the terms of the model. However, certain assumptions about the data are necessary: a normal distribution and constant variance of the residuals with the mean response. Therefore, GAMs provide a more suitable statistical method to analyze biomedical longitudinal data, when these assumptions of the model are met by the data.

However, it is possible that the assumptions of GAMs for longitudinal data do not hold under  certain circumstances. In that case, the field of _Bayesian statistics_ represents a relatively new area of Statistics that does not rely on _p-values_ and hypothesis tests to analyze information. Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine the outcome in regard to significance and are able to expand the number of comparisons and inferences derived form the analysis. On the other hand, the shift that Bayesian statistics represent from the traditional "frequentist" statistical view in research, the computational tools required for its implementation, and the underlying mathematical theory have limited the use of this approach in the biomedical research community. However,Bayesian theory is intuitive and shares some principles with "frequentist" statistics, and there is an increasing use and recognition of the advantages of their use across different areas of biomedical research such as clinical trial design and imaging[@biswas2009;@kelter2020;@kwon2020;@zhou2017].

@@

\begin{equation}
(\#eq:labelANOVA)
y_{hij}=\mu+\gamma_h+\tau_j+(\gamma \tau)_{hj}+\pi_{i(h)}+e_{hij}
\end{equation}

@@

$i=1,...,n_h$ $j=1,...,t$, $h=1,...,p$; with $\pi_i{h} \sim N(0,\sigma^{2}_\epsilon)$ (independently normally distributed) and $e_{hij} \sim N(0,\sigma^{2}_\epsilon)$

@@

In a longitudinal study, the fact that multiple measures are taken on the same subject creates a a _covariance_ issue that needs to be incorporated into the model. Specifically, a parametric model (rm-ANOVA or LMEMs) assumes that there is no relationship between the repeated measures of different subjects (no correlation), but that there is a relationship for the repeated measures of the same subject. The latter is then taken in to account by a _covariance_ structure [@wolfinger1996].  In this case, _covariance_ can be defined as the dependency  between two different values, e.g. if higher values of a variable correspond with higher values of another value, the covariance is positive.  Although not immediately apparent, the reason for this specification arises from the fact that the model from \@ref(eq:labelANOVA) is re-written in matrix form for computational purposes (For an in-depth analysis see [@west2014;@weiss2005]).

@@

In a longitudinal study, missing observations are an important issue that arises frequently. In biomedical research, this situation is not unusual and can be caused by different reasons in the context of each study, but they are typically beyond the control of the investigator [molenberghs2004].statistical treatment of missing observations is particularly important for longitudinal studies. 
*talk about missing at random, mcar etc*

One of the key assumptions for rm-ANOVA is the constant correlation among measurements. This frequently not the case as the correlation reduces as the time interval between two measurements increases []. From a practical

>>>>

A longitudinal study is defined as that which is designed to repeatedly measure a variable of interest in a group (or groups) of subjects. In biomedical research, this type of study arises when the investigator intends to observe the evolution of the effect of a certain treatment across time, rather than analyzing it at a single time point (a cross-sectional study). Clinical examples of this  approach in biomedical research include studies on breast and neck cancer [@sio2016;@kamstra2015]; in the first case, weekly measurements of skin toxicities in  patients with radiation-induced dermatitis were taken for up to 8 weeks; whereas in the latter mouth opening was assessed at 6,12, 18, 24 and 36 months after radiotherapy (RT). Longitudinal studies have used also to measure tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010]. From a statistical standpoint, a longitudinal study presents advantages over a cross-sectional approach: it requires a lower number of subjects to reach a certain statistical power, and besides it being able to track the previously mentioned time-effect evolution on a group-by-group basis, it allows to determine the variability of the response within subjects [@guo2013; @fitzmaurice2012]. In other words,a longitudinal study permits to quantify how the variable changes within each subject across time.

@@

However, constant correlation is frequently unjustified as its value tends to diminish between measures when the time interval between them increases [@ugrinowitsch2004], and the violation of this assumption increases the false positivity rate [@lane2016].
Moreover, it is unlikely that complete observations in all subjects are obtained in a biomedical study due to reasons that can be specific to different situations. In a clinical trial, voluntary withdrawal from one or multiple patients can occur, whereas attrition in animals due to injury or weight loss can occur in preclinical experiments, and in both cases it is possible that unexpected complications with equipment or supplies arise, preventing the researcher from collecting measurements at a certain time point.

When the aforementioned issues arise, rm-ANOVA requires exclusion of all subjects with missing observations from the analysis, thereby increasing costs for the study if the desired statistical power is not met with the remaining observations as it makes necessary to enroll more subjects; and  raising the possibility that the rejection of those partial observations limits the demonstration of significant differences between groups. Additionally, rm-ANOVA uses a _post hoc_ analysis to assess the relevance between the measured response in different groups. This analysis is based on multiple repeated comparisons to estimate a _p-value_, a metric that is widely used as a measure of significance. Because the p-value is highly variable [@halsey2015;@nuzzo2014], multiple comparisons can inflate the false positivity rate [@liu2010],consequently biasing the conclusions of the study.

@@

Additionally, another assumption made in both rm-ANOVA and LMEMs models is that a linear relationship is expected between the observed response and the covariates across the study [@pinheiro2006].This common assumption to both rm-ANOVA and LMEMs causes the models to be restrictive in their inferences when used in longitudinal data that does not follow a linear trend. In biomedical research, this particular behavior in longitudinal has been reported in studies of tumor response to radio/chemotherapy in preclinical and clinical settings[@vishwanath2009; @roblyer2011;@tank2020;@skala2010;@demidov2018], and wound healing and metabolism[@jones2018;@grice2010;@young1994]. These studies have shown that that the collected signal does not follow a linear trend over time, and presents high variability at different time points, making the estimations of a LMEM or rm-ANOVA model inconsistent with the pattern of the observed variations. Additionally,  although it is possible that a post hoc analysis is able to find “significance”( _p-value_<0.05) by using multiple comparisons between the model terms, this estimator would be inherently biased because of the lack of fit between the information and the model.

As  the “frequentist” rm-ANOVA and the more advanced LMEM approach are both limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumed distribution) to determine the fit of the model while enabling inferences that are both adequate and consistent from a statistical perspective.

Generalized additive models (GAMs) are a subset of generalized linear models that use _smooth functions_ (hencefort _smooths_) to estimate the parameters of a model. They have been used in palaeolimnology, ecology and clinical studies to model longitudinal data [@woolway2016;@hefley2017;@ko2007]. Briefly, GAMs use a combination of multiple functions (basis functions) to construct the smooths of the model [@wood2017]. Their main advantage over LMEMs and rm-ANOVA is that the model specification is directed by the _smooths_ rather than by a parametric relationship. This allows a consistent fit of the model with the data, and estimations of significance using the terms of the model. However, certain assumptions about the data are necessary: a normal distribution and constant variance of the residuals with the mean response. Therefore, GAMs provide a more suitable statistical method to analyze biomedical longitudinal data, when these assumptions of the model are met by the data.

However, it is possible that the assumptions of GAMs for longitudinal data do not hold under  certain circumstances. In that case, the field of _Bayesian statistics_ represents a relatively new area of Statistics that does not rely on _p-values_ and hypothesis tests to analyze information. Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine significant differences and are able to expand the number of comparisons and inferences derived form the analysis. On the other hand, the shift that Bayesian statistics represent from the traditional "frequentist" statistical view in research, the computational tools required for its implementation, and the underlying mathematical theory have limited the use of this approach in the biomedical research community. However,Bayesian theory is intuitive and shares some principles with "frequentist" statistics, and there is an increasing use and recognition of the advantages of their use across different areas of biomedical research such as clinical trial design and imaging[@biswas2009;@kelter2020;@kwon2020;@zhou2017].

@@

\begin{equation}
(\#eq:labelANOVA)
y_{kit}=\mu+\gamma_k+\tau_t+(\gamma \tau)_{kt}+\pi_{i(k)}+e_{kit}
\end{equation}

@@

$i=1,...,n_K$ $t=1,...,T$, $k=1,...,K$; with $\pi_i{h} \sim N(0,\sigma^{2}_\epsilon)$ (independently normally distributed) and $e_{hij} \sim N(0,\sigma^{2}_\epsilon)$

@@

The fact that multiple measures are taken on the same subject creates a a _covariance_ issue that needs to be incorporated into the model. Specifically, a parametric model (rm-ANOVA or LMEMs) assumes that there is no relationship between the repeated measures of different subjects (no correlation), but that there is a relationship for the repeated measures of the same subject. The latter is then taken in to account by a _covariance_ structure [@wolfinger1996].  In this case, _covariance_ can be defined as the dependency  between two different values, e.g. if higher values of a variable correspond with higher values of another value, the covariance is positive.  Although not immediately apparent, the reason for this specification arises from the fact that the model from \@ref(eq:labelANOVA) is re-written in matrix form for computational purposes (For an in-depth analysis see [@west2014;@weiss2005]).

@@

In a longitudinal study, missing observations are an issue that arises frequently. In biomedical research, this situation can be caused by reasons beyond the control of the investigator [molenberghs2004].Dropout from patients, or attrition or injury in animals are among the reasons for missing observations. Statistically, missing information can be classified as _missing at random_ (MAR), _missing completely at random_ (MCAR), and _missing not at random_ (MNAR) [@weiss2005].  In a MAR scenario, the pattern of the missing information is related to some variable in the dataset, but it is not related to the variable of interest [@scheffer2002]. If the data are MCAR, this means that the missigness is completely unrelated to the collected information[@potthoff2006], and in the case of MNAR the missing values are dependent on their value. In the case of LMEMs, one key advantage over rm-ANOVA is that the model can work with missing observations that are MAR, whereas rm-ANOVA needs to

*talk about missing at random, mcar etc*
